Explorations to Optimize the Parareal Algorithm \\to Solve ODEs in Given Applications

### 2014-12-09 paper.tex
minor changes on paper
style modifications to paper

%        File: paper.tex
%     Created: Sat Dec 06 04:00 PM 2014 E
%
\documentclass[letterpaper,twocolumn,11pt]{article}
% Change the header if you're going to change this.

% Possible packages - uncomment to use
\usepackage{amsmath}       % Needed for math stuff
%\usepackage{lastpage}      % Finds last page
\usepackage{amssymb}       % Needed for some math symbols
\usepackage{graphicx}      % Needed for graphics
\usepackage[usenames,dvipsnames]{xcolor} % Needed for graphics and color
%\usepackage{setspace}      % Needed to set single/double spacing
\usepackage{float}         % Better placement of figures & tables
\usepackage{hyperref}           % Can have actual links
\usepackage{mathpazo}
\usepackage[linesnumbered,vlined,ruled]{algorithm2e} % algorithm

@@

% Put your stuff here
\twocolumn[
    \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
    We implement the parareal algorithm to solve various ODEs and compare in
    terms of accuracy, speedup and efficiency. We explore the effect of using
    differnt solution operators (a coarse and a fine as required by the method).
    Our code is written in python using mpi4py to parallelize. We observe
    slight speedups in our small scale tests up to 64 processors on Harvard's
    Odyssey cluster. We analyze possible reasons as to why our implementation
    may not be ideal and some tradeoffs that are taken in the parareal
    implementation. Further optimizations are proposed and rationalized as next
    steps including ports of to C++ and BLAS libraries which is a more
    performance-oriented, lower-level language than Python. Data is shown for
    solving for variuos complexities of ODEs from the basic exponential to
    modelling sound wave propagation as we did in Homework 4.
    \end{abstract}
    \end{@twocolumnfalse}
]
\section{Background}

\subsection{Numerical Methods and \\ Parallelization}
Solving systems of differential equations can be a computationally expensive
task. The error of most algorithms scales on the stepsize of the discretization
of time. However, stepsize in time is also proportional to the computation
required. It would be nice to allow for strong scaling where a problem can be
solved in a reasonable time for very small time steps.

Another obstacle to trying to paralleliize numerical methods is that the methods are inherently serial in time - in that evaluation of the next time $n+1$ depends on the previous values, say those at time $n-1$ and $n$.  This setup does not allow for parallelism.  The only way to incorporate parallelism is to create schemes that have parallelizable components - such as an update or a refinement to a serial portion.  As will be discussed and analyzed in the metodology section, one such scheme would be the parareal algorithm.  For an overview, the parareal algorithm first does a coarse (low-order and fast) approximation, then seeks to iteratively correct using smaller, more refined numerical methods done in parallel from approximated starting points approximated by the coarse solution.

An alternative paradigm to applying parallelism to numerical methods is the parallelism by space paradigm, where the result vector space is divided up to different regions and sent to different responsible processors.  In this way, each processor sees a similar but smaller-scale problem - weak scaling.  This method is expected to outperform the parallel in time paradigm simply because the task is somewhat embarassingly parallel, being able to be naturally divided into smaller but identical in method problems.  There is no serial component in this algorithm and should be exhibit both strong and weak scaling.

In general, the division in space paradigm is more intuitive to parallelize and will often lead to greater speedups because there is no serial component.  Yet it is worth noting that knowing how to divide the result vector requires knowledge of problem-specific details.  Thus, this approach lacks the generalisability and theoretical derivaions that schemes that parallelize in time can offer.

@@

The parareal algorithm, developed by Lions, Maday and Turinici in 2001, CITE is
a general algorithm that allows parallelization in time.  It does so by using a
cheaper (lower order, lesser resolution) approximation first, and then makes
corrections in parallel. The entire process is then iterative and tuned for a
given amount of iterations. Therefore, the algorithm is $k$ repetitions of a
serial coarse method updated by a finer method run in parallel for subsections.

\begin{algorithm}[t]
    \KwIn{Temporal discretization $t^n = t_0 + n \Delta t, \, n = 1,2,\ldots,N$}
    \KwIn{Coarse scheme $g_{\Delta t}$}
    \KwIn{Finer	 scheme $g_{\textnormal{fine}}$}
    Compute $u^1_{n+1} = g_{\Delta t}(t_n, u^1_n)$\;
    Compute the corrections $\delta g_n(u^1_n) = g_{\textnormal{fine}}(t_n,
    u^1_n) - g_{\Delta t}(t_n, u^1_n)$ in parallel\;
    Add the prediction and correction terms as $u^2_{n+1} = g_{\Delta t}(t_n,
    u^2_n) + \delta g_n(u^1_n)$\;
    Repeat steps 2 and 3, incrementing the iteration label and using $u^{k+1}_0
    = u^1_0$ as the initial condition\;
 \caption{Parareal}
 \label{alg:parareal}
\end{algorithm}

@@

Show that this parareal method will approach, with large enough k (enough iterations) to approach the error of the fine method.  However, with too large of a k and a lower quality factor (ratio of fine to coarse), the time for the parareal could potentially take longer than the direct serial computation.

@@

We looked at both $e^x$ as well as TODO WHAT OTHER SCHEME using GET PARAMETERES

@@

TODO bsim and ashi show graphs and some times for the forward euler parareal versions - the fastest ones or both if we have obht implemented

@@

On the other hand, paralleism by space already requires a very problem-specific
design. The strong scaling efficiency of parallelism by space is high and is
conceptually embarrisingly parallel and easier to create. However, the
limitations are in the requirement for the code to be designed for the
problem---how to divide by space so as to allow for the maximum number of even
divisions. The lack of generalism is not as beautiful as the parareal algorithm
but may lend itself to faster speedups.

>>>>

%ssssFile: paper.tex
%ss Created: Sat Dec 06 04:00 PM 2014 E
%
\documentclass[letterpaper,12pt]{article}
% Change the header if you're going to change this.

% Possible packages - uncomment to use
\usepackage{amsmath} % Needed for math stuff
%\usepackage{lastpage}sss% Finds last page
\usepackage{amssymb} % Needed for some math symbols
\usepackage{graphicx}% Needed for graphics
\usepackage[usenames,dvipsnames]{xcolor} % Needed for graphics and color
%\usepackage{setspace}sss% Needed to set single/double spacing
\usepackage{float} % Better placement of figures & tables
\usepackage{hyperref} % Can have actual links
\usepackage{mathpazo}
\usepackage[linesnumbered,vlined,ruled]{algorithm2e} % algorithm

@@

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
We implement the parareal algorithm to solve various ODEs and compare in
terms of accuracy, speedup and efficiency. We explore the effect of using
differnt solution operators (a coarse and a fine as required by the method).
Our code is written in python using mpi4py to parallelize. We observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ and BLAS libraries which is a more
performance-oriented, lower-level language than Python. Data is shown for
solving for variuos complexities of ODEs from the basic exponential to
modelling sound wave propagation as we did in Homework 4.
\end{abstract}
%\end{@twocolumnfalse}
%]

\clearpage

\tableofcontents

\clearpage

\section{Background}

\subsection{Numerical Methods and Parallelization}
Solving systems of differential equations can be a computationally expensive
task. The error of most algorithms scales on the stepsize of the discretization
of time. However, stepsize in time is also proportional to the computation
required. It would be nice to allow for strong scaling where a problem can be
solved in a reasonable time for very small time steps.

Another obstacle to trying to paralleliize numerical methods is that the methods
are inherently serial in time---in that evaluation of the next time $n+1$
depends on the previous values, say those at time $n-1$ and $n$. This setup
does not allow for parallelism. The only way to incorporate parallelism is to
create schemes that have parallelizable components---such as an update or a
refinement to a serial portion. As will be discussed and analyzed in the
metodology section, one such scheme would be the parareal algorithm.sFor an
overview, the parareal algorithm first does a coarse (low-order and fast)
approximation, then seeks to iteratively correct using smaller, more refined
numerical methods done in parallel from approximated starting points
approximated by the coarse solution.

An alternative paradigm to applying parallelism to numerical methods is the
parallelism by space paradigm, where the result vector space is divided up to
different regions and sent to different responsible processors. In this way,
each processor sees a similar but smaller-scale problem---weak scaling. This
method is expected to outperform the parallel in time paradigm simply because
the task is somewhat embarassingly parallel, being able to be naturally divided
into smaller but identical in method problems. There is no serial component in
this algorithm and should be exhibit both strong and weak scaling.

In general, the division in space paradigm is more intuitive to parallelize and
will often lead to greater speedups because there is no serial component. Yet
it is worth noting that knowing how to divide the result vector requires
knowledge of problem-specific details. Thus, this approach lacks the
general stability and theoretical derivations that schemes that parallelize in
time can offer.

@@

The parareal algorithm, developed by Lions, Maday and Turinici in 2001, CITE is
a general algorithm that allows parallelization in time.sIt does so by using a
cheaper (lower order, lesser resolution) approximation first, and then makes
corrections in parallel. The entire process is then iterative and tuned for a
given amount of iterations. Therefore, the algorithm is $k$ repetitions of a
serial coarse method updated by a finer method run in parallel for subsections.

\begin{algorithm}[t]
    \KwIn{Temporal discretization $t_n = t_0 + n \Delta t, \, n = 1,2,\ldots,N$}
    \KwIn{Coarse scheme $g_{\Delta t}$}
    \KwIn{Finersscheme $g_{\textnormal{fine}}$}
    Compute $u^1_{n+1} = g_{\Delta t}(t_n, u^1_n)$\;
    Compute the corrections $\delta g_n(u^1_n) = g_{\textnormal{fine}}(t_n,
    u^1_n) - g_{\Delta t}(t_n, u^1_n)$ in parallel\;
    Add the prediction and correction terms as $u^2_{n+1} = g_{\Delta t}(t_n,
    u^2_n) + \delta g_n(u^1_n)$\;
    Repeat steps 2 and 3, incrementing the iteration label and using $u^{k+1}_0
    = u^1_0$ as the initial condition\;
 \caption{Parareal}
 \label{alg:parareal}
\end{algorithm}

@@

Show that this parareal method will approach, with large enough $k$ (enough
iterations) to approach the error of the fine method. However, with too large of
a $k$ and a lower quality factor (ratio of fine to coarse), the time for the
parareal could potentially take longer than the direct serial computation.

@@

We looked at both $e^x$ as well as TODO WHAT OTHER SCHEME using GET PARAMETERS

@@

TODO bsim and ashi show graphs and some times for the forward Euler parareal
versions---the fastest ones or both if we have both implemented

@@

On the other hand, parallelism by space already requires a very problem-specific
design. The strong scaling efficiency of parallelism by space is high and is
conceptually embarrassingly parallel and easier to create. However, the
limitations are in the requirement for the code to be designed for the
problem---how to divide by space so as to allow for the maximum number of even
divisions. The lack of generality is not as beautiful as the parareal algorithm
but may lend itself to faster speedups.
